---
title: "mfinn13_OriginalHomeworkCode_03"
output: html_document
date: "2025-03-13"
---

*Loading in the data set*
```{r}
library(curl)
file <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/refs/heads/master/AN588_Spring25/zombies.csv")
zombies <- read.csv(file, header = T, sep = ",")
```

# Problem 1 
Calculate the population mean and standard deviation for each quantitative random variable (height, weight, age, number of zombies killed, and years of education). NOTE: You will not want to use the built in var() and sd() commands as these are for samples.
```{r}
#viewing dataset so I know the variable names
View(zombies)

#function for calculating population standard deviation
pop_SD <- function(data) {
  sqrt(sum((data - mean(data))^2)/(length(data)))
}

#height
print(mean(zombies$height))
print(pop_SD(zombies$height))

#weight
print(mean(zombies$weight))
print(pop_SD(zombies$weight))

#age
print(mean(zombies$age))
print(pop_SD(zombies$age))

#number of zombies killed
print(mean(zombies$zombies_killed))
print(pop_SD(zombies$zombies_killed))

#years of education
print(mean(zombies$years_of_education))
print(pop_SD(zombies$years_of_education))

```

# Problem 2
Use {ggplot} to make boxplots of each of these variables by gender.
```{r}
library(ggplot2)

#height
heightVgender_Box <- ggplot(data = zombies, aes(x = gender, y = height, fill = gender)) + geom_boxplot()
print(heightVgender_Box)

#weight
weightVgender_Box <- ggplot(data = zombies, aes(x = gender, y = weight, fill = gender)) + geom_boxplot()
print(weightVgender_Box)

#age
ageVgender_Box <- ggplot(data = zombies, aes(x = gender, y = age, fill = gender)) + geom_boxplot()
print(ageVgender_Box)

#number of zombies killed
killsVgender_Box <- ggplot(data = zombies, aes(x = gender, y = zombies_killed, fill = gender)) + geom_boxplot()
print(killsVgender_Box)

#years of education 
educationVgender_Box <- ggplot(data = zombies, aes(x = gender, y = years_of_education, fill = gender)) + geom_boxplot()
print(educationVgender_Box)
```

# Problem 3
Use {ggplot} to make scatter plots of height and weight in relation to age. Do these variables seem to be related? In what way?
```{r}
#height and age relation scatterplot
heightAge_scatter <- ggplot(data = zombies, aes(x = age, y = height)) + geom_point() + geom_smooth(method = "lm")
heightAge_scatter

#weight and age relation scatterplot
weightAge_scatter <- ggplot(data = zombies, aes(x = age, y = weight)) + geom_point() + geom_smooth(method = "lm")
weightAge_scatter

print("Both height and weight seem to be positively correlated with age, though height seems to have a stronger correlation.")



```

# Problem 4
Using histograms and Q-Q plots, check whether the quantitative variables seem to be drawn from a normal distribution. Which seem to be and which do not (hint: not all are drawn from the normal distribution)? For those that are not normal, can you determine from which common distribution they are drawn?
```{r}
#height
hist(zombies$height)
qqnorm(zombies$height)
qqline(zombies$height, col = "blue")
print("Based on the histogram and the Q-Q plot, height seems to be normally distributed.")

#weight
hist(zombies$weight)
qqnorm(zombies$weight)
qqline(zombies$weight, col = "blue")
print("Based on the histogram and the Q-Q plot, weight seems to be normally distributed.")

#age
hist(zombies$age)
qqnorm(zombies$age)
qqline(zombies$age, col = "blue")
print("Based on the histogram and the Q-Q plot, age seems to be normally distributed.")

#zombie kills
hist(zombies$zombies_killed)
qqnorm(zombies$zombies_killed)
qqline(zombies$zombies_killed, col = "blue")
print("The histogram is heavily right-skewed, and the Q-Q plot does not fall on the line. Zombies killed is not normally distributed. Because the number of kills is count data which could take on a value between 0 and infinity, this is likely from a poisson distribution.")

#years of education 
hist(zombies$years_of_education)
qqnorm(zombies$years_of_education)
qqline(zombies$years_of_education, col = "blue")
print("The histogram is right-skewed, and the Q-Q plot does not fall on the line. Years of education is not normally distributed. Because the number of kills is count data which could take on a value between 0 and infinity, this is likely also from a poisson distribution.")


```

# Problem 5
Now use the sample() function to sample ONE subset of 30 zombie survivors (without replacement) from this population and calculate the mean and sample standard deviation for each variable. Also estimate the standard error for each variable, and construct the 95% confidence interval for each mean. Note that for the variables that are not drawn from the normal distribution, you may need to base your estimate of the CIs on slightly different code than for the normal…
```{r}
#First, create a subset of 30 random survivors
set.seed(1) #so it doesn't change every time
randomSurvivors_30 <- zombies[sample(nrow(zombies), 30), ] #selects 30 random rows out of the dataset, and chooses those rows (which are each survivors with all of their data)
randomSurvivors_30


#HEIGHT (normal)
rs30_mean_height <- mean(randomSurvivors_30$height) #calculates mean height
rs30_sd_height <- sd(randomSurvivors_30$height) #calculates standard deviation (we can use sd() because this is a sample)
rs30_se_height <- rs30_sd_height/sqrt(30) #calculates standard error
qnorm(0.975) #the critical value for a normal distribution (which is symmetrical) when alpha is 0.05 (95% confidence) and the sample is greater than or equal to 30
rs30_CIupper_height <- rs30_mean_height + (rs30_se_height*qnorm(0.975)) #calculates CI upper limit
rs30_CIlower_height <- rs30_mean_height - (rs30_se_height*qnorm(0.975)) #calculates CI lower limit
cat("The mean height is",rs30_mean_height,". The standard deviation for height is",rs30_sd_height, ".The standard error for height is",rs30_se_height, ".The 95% confidence interval for the mean of the height is (",rs30_CIlower_height,",", rs30_CIupper_height,").")

#WEIGHT (normal)
rs30_mean_weight <- mean(randomSurvivors_30$weight)
rs30_sd_weight <- sd(randomSurvivors_30$weight) #calculates standard deviation (we can use sd() because this is a sample)
rs30_se_weight <- rs30_sd_weight/sqrt(30) #calculates standard error
qnorm(0.975) #thethe critical value for a normal distribution (which is symmetrical) when alpha is 0.05 (95% confidence) and the sample is greater than or equal to 30
rs30_CIupper_weight <- rs30_mean_weight + (rs30_se_weight*qnorm(0.975)) #calculates CI upper limit
rs30_CIlower_weight <- rs30_mean_weight - (rs30_se_weight*qnorm(0.975)) #calculates CI lower limit
cat("The mean weight is",rs30_mean_weight,". The standard deviation for weight is",rs30_sd_height, ".The standard error for weight is",rs30_se_weight, ".The 95% confidence interval for the mean of the weight is (",rs30_CIlower_weight,",", rs30_CIupper_weight,").")

#AGE (normal)
rs30_mean_age <- mean(randomSurvivors_30$age)
rs30_sd_age <- sd(randomSurvivors_30$age) #calculates standard deviation (we can use sd() because this is a sample)
rs30_se_age <- rs30_sd_age/sqrt(30) #calculates standard error
qnorm(0.975) #the critical value for a normal distribution (which is symmetrical) when alpha is 0.05 (95% confidence) and the sample is greater than or equal to 30.
rs30_CIupper_age <- rs30_mean_age + (rs30_se_age*qnorm(0.975)) #calculates CI upper limit
rs30_CIlower_age <- rs30_mean_age - (rs30_se_age*qnorm(0.975)) #calculates CI lower limit
cat("The mean age is",rs30_mean_age,". The standard deviation for age is",rs30_sd_age, ".The standard error for age is",rs30_se_age, ".The 95% confidence interval for the mean of the age is (",rs30_CIlower_age,",", rs30_CIupper_age,").")


#ZOMBIE KILLS (non-normal)
rs30_mean_zombies_killed <- mean(randomSurvivors_30$zombies_killed)
rs30_sd_zombies_killed <- sd(randomSurvivors_30$zombies_killed) #calculates standard deviation (we can use sd() because this is a sample)
rs30_se_zombies_killed <- rs30_sd_zombies_killed/sqrt(30) #calculates standard error
#because this is not normally distributed, we have to use bootstrapping to estimate a confidence interval
setOfMeans_kills <- NULL #sets null object which we will full with simulated means
for (i in 1:10000){
  setOfMeans_kills[i] <- mean(sample(zombies$zombies_killed, 30, replace = F)) #creates a dataset of 10000 means, each calculated from a random sample of size 30 from our zombies dataset
}
rs30_CIlower_kills <- quantile(setOfMeans_kills, 0.025) #tells us the value in the set of means which 2.5% of values are lower (this is our lower bound)
rs30_CIupper_kills <- quantile(setOfMeans_kills, 0.975) #tell us the value in the set of means which 97.5% of values are lower (this is our upper bound)
cat("The mean number of kills is",rs30_mean_zombies_killed,". The standard deviation for number of kills is",rs30_sd_zombies_killed, ".The standard error for kills is",rs30_se_zombies_killed, ".The 95% confidence interval for the mean of the number of kills is (",rs30_CIlower_kills,",", rs30_CIupper_kills,").")


#YEARS OF EDUCATION (non-normal)
rs30_mean_years_of_education <- mean(randomSurvivors_30$years_of_education)
rs30_sd_years_of_education <- sd(randomSurvivors_30$years_of_education) #calculates standard deviation (we can use sd() because this is a sample)
rs30_se_years_of_education <- rs30_sd_years_of_education/sqrt(30) #calculates standard error
#because this is not normally distributed, we have to use bootstrapping to estimate a confidence interval
setOfMeans_education <- NULL #sets null object which we will full with simulated means
for (i in 1:10000){
  setOfMeans_education[i] <- mean(sample(zombies$years_of_education, 30, replace = F)) #creates a dataset of 10000 means, each calculated from a random sample of size 30 from our zombies dataset
}
rs30_CIlower_education <- quantile(setOfMeans_education, 0.025) #tells us the value in the set of means which 2.5% of values are lower (this is our lower bound)
rs30_CIupper_education <- quantile(setOfMeans_education, 0.975) #tell us the value in the set of means which 97.5% of values are lower (this is our upper bound)
cat("The mean years of education is",rs30_mean_years_of_education,". The standard deviation for years of education is",rs30_sd_years_of_education, ".The standard error for years of education is",rs30_se_years_of_education, ".The 95% confidence interval for the mean of years of education is (",rs30_CIlower_education,",", rs30_CIupper_education,").")

```

# Problem 6
Now draw 99 more random samples of 30 zombie apocalypse survivors, and calculate the mean for each variable for each of these samples. Together with the first sample you drew, you now have a set of 100 means for each variable (each based on 30 observations), which constitutes a sampling distribution for each variable. What are the means and standard deviations of this distribution of means for each variable? How do the standard deviations of means compare to the standard errors estimated in [5]? What do these sampling distributions look like (a graph might help here)? Are they normally distributed? What about for those variables that you concluded were not originally drawn from a normal distribution?
```{r}

#HEIGHT
sampleMeans_height <- NULL #creating dataset which will be filled
for (i in 1:100){
  sampleMeans_height[i] <- mean(sample(zombies$height, 30, replace = F)) #this for loop fills up sampleMeans_height with 100 mean heights calculated from samples of size 30 from the zombies dataset. 
}
mean_sampleMeans_height <- mean(sampleMeans_height) #calculates mean
sd_sampleMeans_height <- sd(sampleMeans_height) #calculates sd
hist(sampleMeans_height) #histogram for assessing normality visually
cat("The mean for the distribution of means of height is", mean_sampleMeans_height, ". The sd for the distribution of means of height is", sd_sampleMeans_height,". while the SE from part 5 was",rs30_se_height, ". The distribution of means for height is normally distributed.")


#WEIGHT
sampleMeans_weight <- NULL #creating dataset which will be filled
for (i in 1:100){
  sampleMeans_weight[i] <- mean(sample(zombies$weight, 30, replace = F)) #this for loop fills up sampleMeans_height with 100 mean heights calculated from samples of size 30 from the zombies dataset. 
}
mean_sampleMeans_weight <- mean(sampleMeans_weight) #calculates mean
sd_sampleMeans_weight <- sd(sampleMeans_weight) #calculates sd
hist(sampleMeans_weight) #histogram for assessing normality visually
cat("The mean for the distribution of means of weight is", mean_sampleMeans_weight, ". The sd for the distribution of means of weight is", sd_sampleMeans_weight,". while the SE from part 5 was",rs30_se_weight, ". The distribution of means for weight is approximately normally distributed.")


#AGE
sampleMeans_age <- NULL #creating dataset which will be filled
for (i in 1:100){
  sampleMeans_age[i] <- mean(sample(zombies$age, 30, replace = F)) #this for loop fills up sampleMeans_height with 100 mean heights calculated from samples of size 30 from the zombies dataset. 
}
mean_sampleMeans_age <- mean(sampleMeans_age) #calculates mean
sd_sampleMeans_age <- sd(sampleMeans_age) #calculates sd
hist(sampleMeans_age) #histogram for assessing normality visually
cat("The mean for the distribution of means of age is", mean_sampleMeans_age, ". The sd for the distribution of means of age is", sd_sampleMeans_age,". while the SE from part 5 was",rs30_se_age, ". The distribution of means for age is approximately normally distributed.")


#ZOMBIE KILLS
sampleMeans_kills <- NULL #creating dataset which will be filled
for (i in 1:100){
  sampleMeans_kills[i] <- mean(sample(zombies$zombies_killed, 30, replace = F)) #this for loop fills up sampleMeans_height with 100 mean heights calculated from samples of size 30 from the zombies dataset. 
}
mean_sampleMeans_kills <- mean(sampleMeans_kills) #calculates mean
sd_sampleMeans_kills <- sd(sampleMeans_kills) #calculates sd
hist(sampleMeans_kills) #histogram for assessing normality visually

cat("The mean for the distribution of means of zombie kills is", mean_sampleMeans_kills, ". The sd for the distribution of means of zombie kills is", sd_sampleMeans_kills,". while the SE from part 5 was",rs30_se_zombies_killed, ". The distribution of means for zombies killed is approximately normally distributed, even though the original distribution was not normally distributed, demonstrating the central limit theorem.")
  

#YEARS OF EDUCATION 
sampleMeans_education <- NULL #creating dataset which will be filled
for (i in 1:100){
  sampleMeans_education[i] <- mean(sample(zombies$years_of_education, 30, replace = F)) #this for loop fills up sampleMeans_height with 100 mean heights calculated from samples of size 30 from the zombies dataset. 
}
mean_sampleMeans_education <- mean(sampleMeans_education) #calculates mean
sd_sampleMeans_education <- sd(sampleMeans_education) #calculates sd
hist(sampleMeans_education) #histogram for assessing normality visually

cat("The mean for the distribution of means of years of education is", mean_sampleMeans_education, ". The sd for the distribution of means of years of education is", sd_sampleMeans_education,". while the SE from part 5 was",rs30_se_years_of_education, ". The distribution of means for years of education is approximately normally distributed, even though the original distribution was not normally distributed, demonstrating the central limit theorem.")



```
# Challenges
1. printing/concatenating characters with numerics, but i found the cat function!
1. problem 5, couldn't get the sample function to work properly with the dataset. 
3. don't fully understand why we would replace when bootstrapping 