---
title: "mfinn13_OriginalHomeworkCode_03"
output: html_document
date: "2025-03-13"
---

Loading in the data set
```{r}
library(curl)
file <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/refs/heads/master/AN588_Spring25/zombies.csv")
zombies <- read.csv(file, header = T, sep = ",")
```

# Problem 1 
Calculate the population mean and standard deviation for each quantitative random variable (height, weight, age, number of zombies killed, and years of education). NOTE: You will not want to use the built in var() and sd() commands as these are for samples.
```{r}
#viewing dataset so I know the variable names
View(zombies)

#function for calculating population standard deviation
pop_SD <- function(data) {
  sqrt(sum((data - mean(data))^2)/(length(data)))
}

#height
print(mean(zombies$height))
print(pop_SD(zombies$height))

#weight
print(mean(zombies$weight))
print(pop_SD(zombies$weight))

#age
print(mean(zombies$age))
print(pop_SD(zombies$age))

#number of zombies killed
print(mean(zombies$zombies_killed))
print(pop_SD(zombies$zombies_killed))

#years of education
print(mean(zombies$years_of_education))
print(pop_SD(zombies$years_of_education))

```

# Problem 2
Use {ggplot} to make boxplots of each of these variables by gender.
```{r}
library(ggplot2)

#height
heightVgender_Box <- ggplot(data = zombies, aes(x = gender, y = height, fill = gender)) + geom_boxplot()
print(heightVgender_Box)

#weight
weightVgender_Box <- ggplot(data = zombies, aes(x = gender, y = weight, fill = gender)) + geom_boxplot()
print(weightVgender_Box)

#age
ageVgender_Box <- ggplot(data = zombies, aes(x = gender, y = age, fill = gender)) + geom_boxplot()
print(ageVgender_Box)

#number of zombies killed
killsVgender_Box <- ggplot(data = zombies, aes(x = gender, y = zombies_killed, fill = gender)) + geom_boxplot()
print(killsVgender_Box)

#years of education 
educationVgender_Box <- ggplot(data = zombies, aes(x = gender, y = years_of_education, fill = gender)) + geom_boxplot()
print(educationVgender_Box)
```

# Problem 3
Use {ggplot} to make scatter plots of height and weight in relation to age. Do these variables seem to be related? In what way?
```{r}
#height and age relation scatterplot
heightAge_scatter <- ggplot(data = zombies, aes(x = age, y = height)) + geom_point() + geom_smooth(method = "lm")
heightAge_scatter

#weight and age relation scatterplot
weightAge_scatter <- ggplot(data = zombies, aes(x = age, y = weight)) + geom_point() + geom_smooth(method = "lm")
weightAge_scatter

print("Both height and weight seem to be positively correlated with age, though height seems to have a stronger correlation.")



```

# Problem 4
Using histograms and Q-Q plots, check whether the quantitative variables seem to be drawn from a normal distribution. Which seem to be and which do not (hint: not all are drawn from the normal distribution)? For those that are not normal, can you determine from which common distribution they are drawn?
```{r}
#height
hist(zombies$height)
qqnorm(zombies$height)
qqline(zombies$height, col = "blue")
print("Based on the histogram and the Q-Q plot, height seems to be normally distributed.")

#weight
hist(zombies$weight)
qqnorm(zombies$weight)
qqline(zombies$weight, col = "blue")
print("Based on the histogram and the Q-Q plot, weight seems to be normally distributed.")

#age
hist(zombies$age)
qqnorm(zombies$age)
qqline(zombies$age, col = "blue")
print("Based on the histogram and the Q-Q plot, age seems to be normally distributed.")

#zombie kills
hist(zombies$zombies_killed)
qqnorm(zombies$zombies_killed)
qqline(zombies$zombies_killed, col = "blue")
print("The histogram is heavily right-skewed, and the Q-Q plot does not fall on the line. Zombies killed is not normally distributed. Because the number of kills is count data which could take on a value between 0 and infinity, this is likely from a poisson distribution.")

#years of education 
hist(zombies$years_of_education)
qqnorm(zombies$years_of_education)
qqline(zombies$years_of_education, col = "blue")
print("The histogram is right-skewed, and the Q-Q plot does not fall on the line. Years of education is not normally distributed. Because the number of kills is count data which could take on a value between 0 and infinity, this is likely also from a poisson distribution.")


```

